\documentclass[10pt,conference,compsocconf]{IEEEtran}
\usepackage{style}
\usepackage{tgpagella}


\begin{document}
\title{
    A Comparative Analysis of Optimization Trajectories and Convergence Properties.
}
\author{
    Ozair Faizan \\
        \texttt{\href{mailto:ahmad.faizan@epfl.ch}{ahmad.faizan@epfl.ch}} \and
    Paul Tercier \\
        \texttt{\href{mailto:paul.tercier@epfl.ch}{paul.tercier@epfl.ch}}\\
}
\maketitle

\setlength{\belowdisplayskip}{2pt} \setlength{\belowdisplayshortskip}{2pt}
\setlength{\abovedisplayskip}{2pt} \setlength{\abovedisplayshortskip}{2pt}


\begin{abstract}
    The choice of the optimization algorithm determines training efficiency and generalization of
    deep neural networks. While methods such as Stochastic Gradient Descent and Adam have become
    field standards, recent works such as Lion \cite{chen2025lionsecretlysolvesconstrained}
    and Muon \cite{jordan2024muon} promise better convergence properties in practice.
    In this paper we present a comparative study of the optimization trajectories produced
    by Stochastic Gradient Descent, Adam, Lion and Muon applied to Multi-Layer
    Perceptron on the Fashion-MNIST dataset \cite{xiao2017/online}.
    To this end, we use a linear interpolation method: for each discrete update
    step from $\theta_t$ to $\theta_{t+1}$, we evaluate the loss function at $N=100?$ points along
    the connecting vector, allowing us to empirically verify convexity and smoothness properties
    along the trajectory.

    Furthermore, we evaluate the convergence points $\vec{m}$ by testing the condition:
    $$ f(\vec{m}) \leqslant f(\vec{x}) + \nabla f(\vec{m})\T (\vec{m} - \vec{x}), $$
    for points $\vec{x}$ within a neighborhood $B_\varepsilon(\vec{m})$. 

    We use constant learning rate found by grid-search and compare them with predicted theoretical
    values, all other hyperparameters are set to default.
\end{abstract}

\section{Introduction}
The choice of the optimizer is central to the process of training deep neural networks, 
influencing not only the speed of convergence but also the ability of the model to generalize. 
Even though there is a vast number of available optimizers that can be applied to train 
a deep neural network, \cite{schneider2021descending} identified over one hundred unique algorithms, 
yet only two have dominated the scene: Stochastic Gradient Descent (SGD) \cite{robbins1951stochastic} and Adam \cite{kingma2015adam}. 
However, recent studies have refuted the common belief. According to \cite{chen2025lionsecretlysolvesconstrained}, 
the Lion optimizer, which was developed based on symbolic search for update equations, 
is able to achieve similar levels of effectiveness without having problems with large amounts of memory usage. 
Furthermore, the Muon optimizer \cite{jordan2024muon} employs Nesterov momentum using a spectral norm-based method.

Empirical comparisons typically report aggregate metrics such as final test accuracy 
or training loss curves. Less studied is the geometry of the optimization trajectory 
itself: whether successive parameter updates traverse locally convex regions of the 
loss landscape, how smooth the loss appears along each step, and whether convergence 
points satisfy local optimality conditions. These properties connect directly to 
quantities central to optimization theory such as Lipschitz smoothness, strong convexity, 
gradient dominance and may explain behavioral differences between optimizers that 
aggregate metrics alone do not reveal.

\section{Models and Methods}
We fit a dense feedforward MLP model on the Fashion MNIST \cite{xiao2017/online}
dataset which includes 10 different classes for image classification.
This dataset contains a total of 60,000 gray-scale images of size 28x28. 
Our model is composed of two layers, each having a width of 256 with activation ReLU, 
followed by a linear classification layer giving rise to a total parameter count of ~270K. 
The training regime includes cross-entropy loss, batch size 128, and 5 epochs. 
No regularization or learning rate decay is employed for simplicity.


The following four optimization algorithms are compared. Stochastic gradient descent (SGD) updates its parameters using
\begin{equation*}
    \theta_{t+1} = \theta_t - \eta g_t.
\end{equation*}
Here, $g_t$ represents the stochastic gradient. The Adam optimizer~\cite{kingma2015adam} computes the first- and second-moment averages of the gradient and performs bias-corrected adaptive steps per coordinate. The Lion optimizer~\cite{chen2025lionsecretlysolvesconstrained} uses only the sign of a momentum-gradient sum as the update direction:
\begin{equation*}
    \theta_{t+1} = \theta_t - \eta \mathrm{sign}\bigl(\beta_1 m_t + (1-\beta_1) g_t\bigr).
\end{equation*}
The momentum is updated by
\begin{equation*}
    m_{t+1} = \beta_2 m_t + (1-\beta_2) g_t.
\end{equation*}
The Muon optimizer~\cite{jordan2024muon} implements the Nesterov-type momentum in the direction of steepest descent in terms of the spectral norm and approximates it by performing the Newton-Schulz orthogonalization of the gradient matrix.

An independent grid search is carried out within the range of learning rates
\begin{equation*}
    \{10^{-4},\, 3\times 10^{-4},\, 10^{-3},\, 3\times 10^{-3},\, 10^{-2}\}.
\end{equation*}
for each optimizer, and the learning rate resulting in the minimum training loss at epoch 50 is chosen. All other hyperparameters remain unchanged from their published values (i.e., $\beta_1$, $\beta_2$ for Adam, Lion, Muon and the momentum coefficient for SGD).

For each optimizer, the parameter vectors $\theta_t$ and $\theta_{t+1}$ are stored at each training iteration. The loss function is calculated at the points along the straight line
\begin{equation*}
    \theta(\alpha) = (1-\alpha)\,\theta_t + \alpha\,\theta_{t+1}, \quad \alpha \in \{0,\, 1/N,\, \ldots,\, 1\},
\end{equation*}
where $N=100$. This analysis allows us to check if the loss function along the update trajectory is convex, concave, or nonmonotone, thus providing a local approximation of the curvature of the loss landscape.

Let $m$ denote the parameter vector at the end of training. We verify the first-order characterization of convexity:
\begin{equation}
    f(m) \leq f(x) + \nabla f(m)^\top (m - x),
    \quad \forall\, x \in \mathcal{B}_\varepsilon(m),
\end{equation}
by sampling $K=50$ directions uniformly on the unit sphere and evaluating $f$ at radius $\varepsilon \in \{10^{-3}, 10^{-2}, 10^{-1}\}$. A convergence point is flagged as satisfying the condition if the inequality holds for all sampled $x$ at a given $\varepsilon$, providing a soft empirical measure of local convexity.

\section{Results}
The selected learning rates and final values for both train loss and test accuracy after 
5 epochs are shown in Table 1. All four optimizers show competitive test accuracy 
on Fashion-MNIST dataset. The highest test accuracy is achieved by both Adam 87.74\% 
and Lion 87.81\%, followed by Muon with 87.29\% and SGD with 85.43\%. Lion optimizer 
achieves an accuracy comparable to Adam with a learning rate that is 10x lower than the 
latter due to the sign-based update producing larger step sizes.

\begin{table}[h]
\centering
\caption{Selected learning rates and final performance after 5 epochs.}
\label{tab:perf}
\begin{tabular}{lcccc}
\hline
Optimizer & LR & Batch Size & Train Loss & Test Acc. (\%) \\
\hline
SGD  & $10^{-2}$ & 32  & 0.3741 & 85.43 \\
Adam & $10^{-3}$ & 64  & 0.2859 & 87.74 \\
Lion & $10^{-4}$ & 128 & 0.2866 & 87.81 \\
Muon & $10^{-3}$ & 64  & 0.3921 & 87.29 \\
\hline
\end{tabular}
\end{table}


\section{Discussion}

\section{Summary}


\newpage

\bibliographystyle{IEEEtran}
\bibliography{literature}

\end{document}


